Thesis: draft

### 2021-05-13 draft.tex
made some small changes, also in appendix

This chapter is subdivided into four main sections. We first discuss the impact of the network structure on the evolution of the prevalence of the opinions and the formation of echo chambers. Then, we introduce some kind of stubbornness in the system. The impact of how the opinions are initially distributed in a network with community structure (SBM) will be discussed afterwards. Finally, we will run the opinion dynamics model of this thesis on some real-world networks. We will discuss the results and compare them to results obtained from network models with a similar average degree, clustering coefficient and/or community structure.

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithm are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has similar clustering and community structure as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering and/or community structure that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution. This agrees with the result that Perra and Rocha \cite{Perra2019} found: a heterogeneous contact pattern hampers the formation of echo chambers. Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The stochastic block model and stochastic Watts-Strogatz block model, that we used, only have communities of a single size. This does, however, not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.\\ % add parameters of SBM-WS power??
\newpage
\noindent
The results for the Watts-Strogatz model are similar to those for the stochastic Watts-Strogatz block model. The formation of echo chambers seems to be slightly smaller in the WS than in the SBM-WS network. It may be tempting to attribute this to the lack of community structure in the WS model. It must however be noted that the implemented WS model has a slightly smaller clustering coefficient ($\left<cc\right> = 0.317 \pm 0.004$) than the SBM-WS ($\left<cc\right> = 0.344 \pm 0.002$), which may also explain the somewhat smaller formation of echo chambers. Furthermore, it is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering. % talk about coincidence??

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease again. The increase is the strongest for the high clustering network models (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. As for the PR filtering algorithm, this seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the corresponding hampering due to the lack of high clustering.

@@

We made use of simple toy models of real social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of several layers, including a network layer and an opinion dynamics layer. Simple filtering algorithms were implemented to replicate and simulate the content curation used by real social media companies. Furthermore, a notion of stubbornness was introduced to investigate the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B).\\

@@

% put here: be carefull with extrapolating data/results to real world cases, need for more real data to compare with (eg data from role of facebook in elections in US), this lack of comparison between real data and network model data/analytical work = one of the downfalls/backsides of social physics??
% relevance to society of knowing the impact of social media on opinion dynamics 
% important to know the limits of the model, may be intuitive, but intuition maybe misleading (learned in physics) --> why we need experimental data to compare with!

@@

Using \cite{Albert2016}
\begin{equation}
	Q_c = \frac{1}{2L}\sum_{(i,j) \in C_c} (A_{ij} - p_{ij}) \ ,
\end{equation}
and \cite{Albert2016}
\begin{equation}
	p_{ij} = \frac{k_ik_j}{2L} \ ,
\end{equation}
we can write the modularity of the whole network as \cite{Albert2016}
\begin{equation}\label{modnetw}
	Q = \frac{1}{2L}\sum_{i, j = 1}^N \bigg(A_{ij} - \frac{k_ik_j}{2L}\bigg) \delta_{c_i, c_j} \ ,
\end{equation}
where $c_i$ represents the community to which node $i$ belongs and where the Dirac-delta function $\delta_{c_i, c_j}$ expresses the fact that only pairs of nodes that belong to the same community are taken into account. We can then rewrite the first term as a sum over communities \cite{Albert2016}
\begin{equation}\label{1}
	\frac{1}{2L}\sum_{i, j = 1}^N A_{ij}\delta_{c_i, c_j} = \sum_{c=1}^{n_c}\frac{1}{2L}\sum_{i, j \in C_c} A_{ij} = \sum_{c=1}^{n_c}\frac{L_c}{L} \ ,
\end{equation}
where $L_c$ is the number of edges in community $C_c$ and the factor 2 disappears because each edge is counted twice in $A_{ij}$. Similarly, the second term in Eq. (\ref{modnetw}) can be rewritten as \cite{Albert2016}
\begin{equation}\label{2}
	\frac{1}{2L}\sum_{i, j = 1}^N \frac{k_ik_j}{2L} \delta_{c_i, c_j} = \sum_{c=1}^{n_c}\frac{1}{(2L)^2}\sum_{i, j \in C_c} k_ik_j = \sum_{c=1}^{n_c}\frac{k_c^2}{4L^2} \ ,
\end{equation}
where $k_c$ is the total degree of the nodes in community $C_c$. Combining Eq. (\ref{1}) and Eq. (\ref{2}) gives the final result \cite{Albert2016}
\begin{equation}\label{Qfin}
	Q = \sum_{c = 1}^{n_c}\bigg[\frac{L_c}{L} - \bigg(\frac{k_c}{2L}\bigg)^2 \bigg] \ .
\end{equation}

@@

Consider two communities $A$ and $B$ and denote with $k_A$ and $k_B$ the total degree of communities $A$ and $B$ respectively \cite{Albert2016}. The change in modularity after merging the two communities is calculated as follows (using Eq. (\ref{Qfin})) \cite{Albert2016}
\begin{equation}\label{Q_AB}
	\Delta Q_{AB} = \bigg[\frac{L_{AB}}{L} - \bigg(\frac{k_{AB}}{2L} \bigg)^2 \bigg] - \bigg[\frac{L_A}{L} - \bigg(\frac{k_A}{2L} \bigg)^2 + \frac{L_B}{2L} - \bigg(\frac{k_B}{2L} \bigg)^2 \bigg] \ ,
\end{equation}
where $L_{AB} = L_A + L_B + l_{AB}$ is the total number of edges in the merged community; $L_A$ and $L_B$ are the number of edges in communities $A$ and $B$ respectively and $l_{AB}$ is the number of direct edges between communities $A$ and $B$ \cite{Albert2016}. $k_{AB} = k_A + k_B$ is the total degree of nodes in the merged community \cite{Albert2016}. Inserting the formulas for $L_{AB}$ and $k_{AB}$ in Eq. (\ref{Q_AB}) gives
\begin{equation}
	\Delta Q_{AB} = \frac{l_{AB}}{L} - \frac{k_Ak_B}{2L^2} \ .
\end{equation}

@@

\subsection{Average degree}\label{degree}
The expected mean degree is given by \cite{Hopcroft2006}
\begin{equation}\label{avDegr}
\begin{split}
	\left<k\right> &= \sum_{k=0}^{N-1} k\  P(k)  \\
&= \sum_{k=0}^{N-1} k \binom{N-1}{k}p^k (1-p)^{N-1-k} \ .
\end{split}
\end{equation}
In order to simplify this equation we can make use of the following formula \cite{Hopcroft2006}
\begin{equation} 
	(p+q)^n = \sum_{k=0}^n \binom{n}{k} p^k q^{n-k} \ .
\end{equation}
Differentiating both sides of this equation (with respect to $p$) gives us \cite{Hopcroft2006}
\begin{equation}
\begin{split}
	n(p+q)^{n-1} &= \sum_{k=0}^n \binom{n}{k} k p^{k-1} q^{n-k} \\
	&= \frac{1}{p} \sum_{k=0}^n \binom{n}{k} k p^{k} q^{n-k} \ .
\end{split}
\end{equation}
Finally, by substituting $q = 1-p$, we get \cite{Hopcroft2006}
\begin{equation}
\begin{split}
	n (p+(1-p))^{n-1} &= \frac{1}{p} \sum_{k=0}^n \binom{n}{k} k p^{k} (1-p)^{n-k} \\
	np &= \sum_{k=0}^n \binom{n}{k} k p^{k} (1-p)^{n-k} \ ,
\end{split}
\end{equation}
where the right hand side is equal to Equation (\ref{avDegr}) with $n = N-1$ and we thus obtain $\left<k\right> = (N-1)p$.

\subsection{Poisson form of the degree distribution}\label{poiss}
We begin with the exact binomial distribution (Eq.(\ref{degDistRan}))
\begin{equation}\label{pk}
	P(k) = \binom{N-1}{k}p^k (1-p)^{N-1-k} \ ,
\end{equation}
which characterizes a random graph. In the case $k \ll N$, we can simplify the binomial in the following way \cite{Albert2014}
\begin{equation}\label{bin}
	\binom{N-1}{k} = \frac{(N-1)(N-1-1)...(N-1-k+1)}{k!} \approx \frac{(N-1)^k}{k!} \ .
\end{equation}
We also note the following equality \cite{Albert2014}
\begin{equation}
	\ln[(1-p)^{N-1-k}] = (N-1-k)\ln\bigg(1 - \frac{\left<k\right>}{N-1} \bigg) \ ,
\end{equation}
where we made use of the fact that $\left<k\right> = (N-1)p$. Using the series expansion \cite{Albert2014}
\begin{equation}
	\ln(1+x) = \sum_{n=1}^{\infty}\frac{(-1)^{n+1}}{n}x^n = x - \frac{x^2}{2} + \frac{x^3}{3} - ... \ ,  \qquad \forall \ |x| \leqslant 1 \ ,
\end{equation}
we get \cite{Albert2014}
\begin{equation}
	\ln[(1-p)^{N-1-k}] \approx (N-1-k) \frac{\left<k\right>}{N-1} = - \left<k\right>\bigg(1 - \frac{\left<k\right>}{N-1} \bigg) \approx - \left<k\right> \ ,
\end{equation}
which is valid if $k \ll N$. We thus find \cite{Albert2014}
\begin{equation}\label{e}
	(1-p)^{N-1-k} = \mathrm{e}^{-\left<k\right>} \ .
\end{equation}
Substituting Eq. (\ref{e}) and Eq. (\ref{bin}) in Eq. (\ref{pk}) and using $\left<k\right> = (N-1)p$, we finally obtain the Poisson form of the degree distribution \cite{Albert2014}
\begin{equation}
\begin{split}
	P(k) &= \binom{N-1}{k}p^k (1-p)^{N-1-k} = \frac{(N-1)^k}{k!} p^k \mathrm{e}^{-\left<k\right>} \\
&= \frac{(N-1)^k}{k!} \bigg(\frac{\left<k\right>}{N-1} \bigg)^k \mathrm{e}^{-\left<k\right>}\\
&= \mathrm{e}^{-\left<k\right>}\frac{\left<k\right>^k}{k!} \ .
\end{split}
\end{equation}

>>>>

This chapter is subdivided into four main sections. We first discuss the impact of the network structure on the evolution of the prevalence of the opinions and the formation of echo chambers. Then, we introduce some kind of stubbornness in the system. The impact of how the opinions are initially distributed in a network with community structure (SBM) will be discussed afterwards. Finally, we will run the opinion dynamics model of this thesis on some real-world networks. We will discuss the results and compare them to results obtained from network models with a similar average degree, clustering coefficient and/or modularity.

@@

The formation of echo chambers versus the fraction of completely stubborn actors, fracRes, in the Last.fm on-line social network is studied. The results are compared to corresponding results from network models, all with $N = 8003$: (i) the Watts-Strogatz model with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) the stochastic block model with 53 communities of size 151, $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$; (iii) the stochastic Watts-Strogatz block model with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$. The results for the PR and REC filtering algorithms are shown in Fig. \ref{echo_vs_fracRes_real}.\\
\newline
The PR filtering algorithm shows a decrease in the formation of echo chambers with an increasing fraction of stubborn actors. This is valid for both the real-world network and the network models. The stochastic Watts-Strogatz block model has similar clustering and modularity as the Last.fm network. The Last.fm network has, however, a smaller formation of echo chambers than the SBM-WS. This suggests that it is not only clustering and/or modularity that drives the formation of echo chambers. Other topologies must play a role as well. The biggest difference between the real-world Last.fm network and the stochastic Watts-Strogatz block model lies in the degree distribution: the Last.fm network has a heterogeneous degree distribution instead of a homogeneous degree distribution. It is probably this heterogeneous contact pattern that hampers the formation of echo chambers in the real-world network compared to a network model with similar clustering and modularity, but with a homogeneous degree distribution. This agrees with the result that Perra and Rocha \cite{Perra2019} found: a heterogeneous contact pattern hampers the formation of echo chambers. Another difference between the Last.fm and the SBM and SBM-WS network is the community structure. The Last.fm network has a community structure where the community sizes follow a heterogeneous distribution: plenty of small communities and a few large ones. The stochastic block model and stochastic Watts-Strogatz block model, that we used, only have communities of a single size. This does, however, not play a significant role in the formation of echo chambers, as can be seen in Fig. \ref{power_vs_single}. In this figure, the distribution of echo chambers in the Last.fm network is compared to those in the SBM-WS network with similar clustering and modularity. The SBM-WS is modeled in two ways: (i) the communities all have a fixed size; (ii) the communities have the same heterogeneous sizes as in the Last.fm network. The PR and REC filtering algorithms are compared.\\ % add parameters of SBM-WS power??
\newpage
\noindent
The results for the Watts-Strogatz model are similar to those for the stochastic Watts-Strogatz block model. The formation of echo chambers seems to be slightly smaller in the WS than in the SBM-WS network. It may be tempting to attribute this to the lack of community structure in the WS model. It must however be noted that the implemented WS model has a slightly smaller clustering coefficient ($\left<cc\right> = 0.317 \pm 0.004$) than the SBM-WS ($\left<cc\right> = 0.344 \pm 0.002$), which may also explain the somewhat smaller formation of echo chambers. Furthermore, it is interesting to notice that the Last.fm on-line social network coincides with the SBM with a similar modularity. It seems that the hampering of the formation of echo chambers due to a heterogeneous contact pattern is of the same size as the hampering due to a lack of high clustering. % talk about coincidence??

\begin{figure}[H]
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{PR}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_PR_real_network_8x7.png}
    \label{pr_real}
  \end{subfigure}
  \begin{subfigure}[b]{0.49\textwidth}
    \caption{REC}
  	\includegraphics[width=8cm, height=7cm]{echo_chamber_vs_fracRes_REC_real_network_8x7.png}
    \label{rec_real}
  \end{subfigure}
  \captionsetup{format=plain}
  \caption[Formation of echo chambers versus fraction of completely stubborn actors for the PR and REC filtering algorithms and an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models.]{Formation of echo chambers versus the fraction of completely stubborn actors, fracRes, for an initial $50/50$ opinion distribution. The Last.fm on-line social network is compared to corresponding network models: (i) WS with $\left<k\right> = 4$, $\beta = 0.15$ and $\left<cc\right> = 0.317 \pm 0.004$; (ii) SBM-WS with 53 communities of size 151, each community has the following Watts-Strogatz parameters: $\left<k\right> = 4$, $\beta = 0.025$; edges are added between communities with probability $p_\text{add} = 0.0001$, $\left<k\right> = 4.77 \pm 0.01$, $\left<cc\right> = 0.344 \pm 0.002$ and $Q = 0.833 \pm 0.003$; (iii) SBM with 53 communities of size 151 with $p_\text{cl} = 0.023$, $p_\text{add} = 0.0001$, $\left<k\right> = 4.24 \pm 0.03$ and $Q = 0.811 \pm 0.004$. \textbf{(a)} PR filtering algorithm; \textbf{(b)} REC filtering algorithm. The $y$-axis shows the average of the normalized fraction of nodes with all neighbors having opinion A and the normalized fraction of nodes with all neighbors having opinion B. This value represents the formation of echo chambers in the system (denoted with echo chamber size in the figure). Each data point is the average of $50$ independent simulations. Bars represent the standard deviation $\sigma$.}
\label{echo_vs_fracRes_real}
\end{figure}
\noindent
The REC filtering algorithm enforces an increase in the formation of echo chambers when a small fraction of stubborn actors is introduced. If this fraction is further increased, the formation of echo chambers starts to decrease. The increase is the strongest for the high clustering network models (WS, SBM-WS). Even though the Last.fm network has a similar clustering as these high clustering models, the formation of echo chambers and the corresponding increase in that formation are significantly smaller. As for the PR filtering algorithm, this is probably a consequence of the heterogeneous degree pattern of the Last.fm network. The results for the Last.fm network coincide with those of the SBM. As for the PR filtering algorithm, this seems to suggest that the hampering in the formation of echo chambers due to a heterogeneous degree distribution is of a similar size as the corresponding hampering due to the lack of high clustering.

@@

We made use of simple toy models of real social media platforms to generate results regarding the evolution of the prevalence of the opinions and formation of echo chambers. These toy models consist of several layers, including a network layer and an opinion dynamics layer. Simple filtering algorithms were implemented to replicate and simulate the content curation used by real social media companies. Furthermore, a notion of stubbornness was introduced to investigate the interplay between individual resistance to change, network structure and filtering algorithm. For simplicity we only allowed two competing opinions (A and B). Furthermore, we studied two different settings: balanced starting conditions and unbalanced starting conditions.\\

@@

@@

Using \cite{Albert2016}
\begin{equation}
	Q_c = \frac{1}{2L}\sum_{(i,j) \in C_c} (A_{ij} - p_{ij}) \ ,
\end{equation}
and \cite{Albert2016}
\begin{equation}
	p_{ij} = \frac{k_ik_j}{2L} \ ,
\end{equation}
we can write the modularity of the whole network as \cite{Albert2016}
\begin{equation}\label{modnetw}
	Q = \frac{1}{2L}\sum_{i, j = 1}^N \bigg(A_{ij} - \frac{k_ik_j}{2L}\bigg) \delta_{c_i, c_j} \ ,
\end{equation}
where $c_i$ represents the community to which node $i$ belongs and the Dirac-delta function $\delta_{c_i, c_j}$ expresses the fact that only pairs of nodes that belong to the same community are taken into account. We can then rewrite the first term as a sum over communities \cite{Albert2016}
\begin{equation}\label{1}
	\frac{1}{2L}\sum_{i, j = 1}^N A_{ij}\delta_{c_i, c_j} = \sum_{c=1}^{n_c}\frac{1}{2L}\sum_{(i, j) \in C_c} A_{ij} = \sum_{c=1}^{n_c}\frac{L_c}{L} \ ,
\end{equation}
where $L_c$ is the number of edges in community $C_c$ and the factor 2 disappears because each edge is counted twice in $A_{ij}$. Similarly, the second term in Eq. (\ref{modnetw}) can be rewritten as \cite{Albert2016}
\begin{equation}\label{2}
	\frac{1}{2L}\sum_{i, j = 1}^N \frac{k_ik_j}{2L} \delta_{c_i, c_j} = \sum_{c=1}^{n_c}\frac{1}{(2L)^2}\sum_{(i, j) \in C_c} k_ik_j = \sum_{c=1}^{n_c}\frac{k_c^2}{4L^2} \ ,
\end{equation}
where $k_c$ is the total degree of the nodes in community $C_c$. Combining Eq. (\ref{1}) and Eq. (\ref{2}) gives the final result \cite{Albert2016}
\begin{equation}\label{Qfin}
	Q = \sum_{c = 1}^{n_c}\bigg[\frac{L_c}{L} - \bigg(\frac{k_c}{2L}\bigg)^2 \bigg] \ .
\end{equation}

@@

Consider two communities $A$ and $B$ and denote with $k_A$ and $k_B$ the total degree of communities $A$ and $B$ respectively \cite{Albert2016}. The change in modularity after merging the two communities is calculated as follows (using Eq. (\ref{Qfin})) \cite{Albert2016}
\begin{equation}\label{Q_AB}
	\Delta Q_{AB} = \bigg[\frac{L_{AB}}{L} - \bigg(\frac{k_{AB}}{2L} \bigg)^2 \bigg] - \bigg[\frac{L_A}{L} - \bigg(\frac{k_A}{2L} \bigg)^2 + \frac{L_B}{L} - \bigg(\frac{k_B}{2L} \bigg)^2 \bigg] \ ,
\end{equation}
where $L_{AB} = L_A + L_B + l_{AB}$ is the total number of edges in the merged community; $L_A$ and $L_B$ are the number of edges in communities $A$ and $B$ respectively and $l_{AB}$ is the number of direct edges between communities $A$ and $B$ \cite{Albert2016}. $k_{AB} = k_A + k_B$ is the total degree of nodes in the merged community; $k_A$ and $k_B$ are the total degree of the nodes in community $A$ and $B$ respectively \cite{Albert2016}. Inserting the formulas for $L_{AB}$ and $k_{AB}$ in Eq. (\ref{Q_AB}) gives
\begin{equation}
	\Delta Q_{AB} = \frac{l_{AB}}{L} - \frac{k_Ak_B}{2L^2} \ .
\end{equation}

@@

\subsection{Average degree}\label{degree}
The expected mean degree is given by \cite{Hopcroft2006}
\begin{equation}\label{avDegr}
\begin{split}
	\left<k\right> &= \sum_{k=0}^{N-1} k\  P(k)  \\
&= \sum_{k=0}^{N-1} k \binom{N-1}{k}p^k (1-p)^{N-1-k} \ .
\end{split}
\end{equation}
In order to simplify this equation we can make use of the following formula \cite{Hopcroft2006}
\begin{equation} 
	(p+q)^n = \sum_{k=0}^n \binom{n}{k} p^k q^{n-k} \ .
\end{equation}
Differentiating both sides of this equation (with respect to $p$) gives us \cite{Hopcroft2006}
\begin{equation}
\begin{split}
	n(p+q)^{n-1} &= \sum_{k=0}^n \binom{n}{k} k p^{k-1} q^{n-k} \\
	&= \frac{1}{p} \sum_{k=0}^n \binom{n}{k} k p^{k} q^{n-k} \ .
\end{split}
\end{equation}
Finally, by substituting $q = 1-p$, we get \cite{Hopcroft2006}
\begin{equation}
\begin{split}
	n (p+(1-p))^{n-1} &= \frac{1}{p} \sum_{k=0}^n \binom{n}{k} k p^{k} (1-p)^{n-k} \\
	np &= \sum_{k=0}^n \binom{n}{k} k p^{k} (1-p)^{n-k} \ ,
\end{split}
\end{equation}
where the right hand side is equal to Eq. (\ref{avDegr}) with $n = N-1$ and we obtain $\left<k\right> = (N-1)p$.

\subsection{Poisson form of the degree distribution}\label{poiss}
We begin with the exact binomial distribution (Eq.(\ref{degDistRan}))
\begin{equation}\label{pk}
	P(k) = \binom{N-1}{k}p^k (1-p)^{N-1-k} \ ,
\end{equation}
which characterizes a random graph. In the case $k \ll N$, we can simplify the binomial in the following way \cite{Albert2014}
\begin{equation}\label{bin}
	\binom{N-1}{k} = \frac{(N-1)(N-1-1)...(N-1-k+1)}{k!} \approx \frac{(N-1)^k}{k!} \ .
\end{equation}
We also note the following equality \cite{Albert2014}
\begin{equation}
	\ln[(1-p)^{N-1-k}] = (N-1-k)\ln\bigg(1 - \frac{\left<k\right>}{N-1} \bigg) \ ,
\end{equation}
where we made use of the fact that $\left<k\right> = (N-1)p$. Using the series expansion \cite{Albert2014}
\begin{equation}
	\ln(1+x) = \sum_{n=1}^{\infty}\frac{(-1)^{n+1}}{n}x^n = x - \frac{x^2}{2} + \frac{x^3}{3} - ... \ ,  \qquad \forall \ |x| \leqslant 1 \ ,
\end{equation}
we get \cite{Albert2014}
\begin{equation}
	\ln[(1-p)^{N-1-k}] \approx -(N-1-k) \frac{\left<k\right>}{N-1} = - \left<k\right>\bigg(1 - \frac{\left<k\right>}{N-1} \bigg) \approx - \left<k\right> \ ,
\end{equation}
which is valid if $k \ll N$. We thus find \cite{Albert2014}
\begin{equation}\label{e}
	(1-p)^{N-1-k} = \mathrm{e}^{-\left<k\right>} \ .
\end{equation}
Substituting Eq. (\ref{e}) and Eq. (\ref{bin}) in Eq. (\ref{pk}) and using $\left<k\right> = (N-1)p$, we finally obtain the Poisson form of the degree distribution \cite{Albert2014}
\begin{equation}
\begin{split}
	P(k) &= \binom{N-1}{k}p^k (1-p)^{N-1-k} = \frac{(N-1)^k}{k!} p^k \mathrm{e}^{-\left<k\right>} \\
&= \frac{(N-1)^k}{k!} \bigg(\frac{\left<k\right>}{N-1} \bigg)^k \mathrm{e}^{-\left<k\right>}\\
&= \mathrm{e}^{-\left<k\right>}\frac{\left<k\right>^k}{k!} \ .
\end{split}
\end{equation}
